---
title: "Subset and label LFS microdata"
author: "Jeremy Buhler"
date: "13/01/2021"
output: html_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

The Labour Force Survey (LFS) from Statistics Canada is released in monthly microdata files. This file processes them, keeping only a small number of variables used in the Tableau workshops offered by the UBC Library Research Commons. In addition to subsetting, many numerical values are replaced with human-readable values.

## Install and load required packages
Uncomment the first line if tidyverse is not already installed on local machine

```{r}
#install.packages("tidyverse")
library(tidyverse)
```


## Import LFS file
In this case the desired file, `LFS_2020_quarterly.tab`, is a union of Jan, Apr, Jul, and Oct LFS microdata. Can be replaced with any LFS microdata file.

```{r}
LFS_data <- read_tsv("../data/LFS_2020_quarterly.tab")
```
## Keep only desired variables
For the Tableau workshops  relatively small number of variables are used

```{r}
LFS_subset <- select(LFS_data,REC_NUM,SURVYEAR,SURVMNTH,LFSSTAT,PROV,CMA,AGE_12,SEX,EDUC,MJH,COWMAIN,FTPTMAIN,NOC_10,UTOTHRS,ATOTHRS,DURUNEMP,WHYPT,WHYLEFTN,HRLYEARN,PERMTEMP,FLOWUNEM,SCHOOLN,FINALWT)
```

## Replace numerical values with labels for selected variables
To save time during the workshop some of the values will be recoded ahead of time. Labels are from the LFS documentation. 

```{r}
LFS_recoded <- LFS_subset %>%
mutate(LFSSTAT=recode(LFSSTAT,
"01" = "Employed, at work",
"02" = "Employed, absent from work",
"03" = "Unemployed",
"04" = "Not in labour force"),
PROV=recode(PROV,
"10" = "Newfoundland and Labrador",
"11" = "Prince Edward Island",
"12" = "Nova Scotia",
"13" = "New Brunswick",
"24" = "Quebec",
"35" = "Ontario",
"46" = "Manitoba",
"47" = "Saskatchewan",
"48" = "Alberta",
"59" = "British Columbia"),
AGE_12=recode(AGE_12,
"01" = "15-19",
"02" = "20-24",
"03" = "25-29",
"04" = "30-34",
"05" = "35-39",
"06" = "40-44",
"07" = "45-49",
"08" = "50-54",
"09" = "55-59",
"10" = "60-64",
"11" = "65-69",
"12" = "70 and over"),
EDUC=recode(EDUC,
"0" = "0 to 8 years",
"1" = "Some high school",
"2" = "High school graduate",
"3" = "Some postsecondary",
"4" = "Postsecondary certificate or diploma",
"5" = "Bachelor's degree",
"6" = "Above bachelor's degree"),
SURVMNTH=recode(SURVMNTH,
"1" = "January",
"2" = "February",
"3" = "March",
"4" = "April",
"5" = "May",
"6" = "June",
"7" = "July",
"8" = "August",
"9" = "September",
"10" = "October",
"11" = "November",
"12" = "December"),
MJH=recode(MJH,
"1" = "Single jobholder",
"2" = "Multiple jobholder"),
COWMAIN=recode(COWMAIN,
"1" = "public sector employees",
"2" = "private sector employees",
"3" = "self-employed incorporated, with paid help",
"4" = "self-employed incorporated, no paid help",
"5" = "self-employed unincorporated, with paid help",
"6" = "self-employed unincorporated, no paid help",
"7" = "unpaid family worker"),
NOC_10=recode(NOC_10,
"01" = "Management",
"02" = "Business, finance and administration",
"03" = "Natural and applied sciences and related",
"04" = "Health",
"05" = "Education, law and social, community and government services",
"06" = "Art, culture, recreation and sport",
"07" = "Sales and service",
"08" = "Trades, transport and equipment operators and related",
"09" = "Natural resources, agriculture and related production",
"10" = "Manufacturing and utilities"),
FTPTMAIN=recode(FTPTMAIN,
"1" = "Full-time",
"2" = "Part-time"),
WHYPT=recode(WHYPT,
"0" = "Other reasons",
"1" = "Own illness or disability",
"2" = "Caring for children",
"3" = "Other personal or family responsibilities",
"4" = "Going to school",
"5" = "Personal preference",
"6" = "Business conditions or could not find full-time work, looked for full-time work in last month",
"7" = "Business conditions or could not find full-time work, did not look for full-time work in last month"),
PERMTEMP=recode(PERMTEMP,
"1" = "Permanent",
"2" = "Temporary, seasonal job",
"3" = "Temporary, term or contract job",
"4" = "Temporary, casual or other temorary jobs"),
WHYLEFTN=recode(WHYLEFTN,
"00" = "Job leavers, other reasons",
"01" = "Job leavers, own illness or disability",
"02" = "Job leavers, caring for children",
"03" = "Job leavers, pregnancy",
"04" = "Job leavers, personal or family responsibilities",
"05" = "Job leavers, going to school",
"06" = "Job leavers, dissatisfied",
"07" = "Job leavers, retired",
"08" = "Job leavers, business sold or closed down (self-employed)",
"09" = "Job losers, end of seasonal job (employee)",
"10" = "Job losers, end of temporary or casual (employee)",
"11" = "Job losers, company moved or out of business (employee)",
"12" = "Job losers, business conditions (employee)",
"13" = "Job losers, dismissal or other reasons"),
SCHOOLN=recode(SCHOOLN,
"1" = "Non-student",
"2" = "Full-time student",
"3" = "Part-time student"),
FLOWUNEM=recode(FLOWUNEM,
"1" = "Job losers, temporary layoff",
"2" = "Job losers, permanent layoff",
"3" = "Job leavers",
"4" = "Job leavers/losers (status unknown), worked more than 1 year ago ",
"5" = "New entrants",
"6" = "Re-entrants, worked 1 year ago or less",
"7" = "Re-entrants, worked more than 1 year ago",
"8" = "Future starts"))
```

## Rename selected variables
To make the more recognizable without reference to the user guide 

```{r}
LFS_renamed <- rename(LFS_recoded, 
       "Labour force status" = LFSSTAT,
       "Highest educational attainment" = EDUC,
       "Census metropolitan area" = CMA,
       "Age group" = AGE_12,
       "Single or multiple jobholder" = MJH,
       "Type of work (main job)" = FTPTMAIN,
       "Class of worker (main job)" = COWMAIN,
       "Occupation at main job" = NOC_10,
       "Usual hours worked/wk at main job" = UTOTHRS,
       "Actual hours worked/wk at main job" = ATOTHRS,
       "Reason for part time work" = WHYPT,
       "Job permanency (employees only)" = PERMTEMP,
       "Usual hourly wages (employees only)" = HRLYEARN,
       "Duration of unemployment" = DURUNEMP,
       "Flows into unemployment" = FLOWUNEM,
       "Student status" = SCHOOLN,
       "Reason for leaving job" = WHYLEFTN,
       "Survey month" = SURVMNTH,
       "Survey year" = SURVYEAR,
       "Record number" = REC_NUM,
       "Province" = PROV,
       "Sex" = SEX,
       "Statistical Weight" = FINALWT)
```



## Save output as csv file
```{r}
write.csv(LFS_renamed,"../LFS_2020_quarterly_recoded_subset.csv", row.names=FALSE, na="")
```

